---
title: "All species CPUE"
author: "Alberto Rovellini"
date: "5/3/2021"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

This document reads in Catch data from AKFIN Answers for all species in the GOA, and calculates the CPUE for each species in each haul based on the Haul Description data set from AKFIN. The CPUE data we calculate this way will be used to run sdmTMB on invertebrate groups. It will be a large data set. 

This is a first version of code used in the SDM routines to convert from AKFIN data to input for `sdmTMB`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(kableExtra)
library(sf)
library(raster)
library(viridis)
library(maps)
library(mapdata)
```

```{r}
select <- dplyr::select
```

Read in Atlantis groups, the RACE species list, and map them to one another.
```{r}
atlantis_groups <- read.csv("../data/GOA_Groups.csv", fileEncoding = "UTF-8-BOM")
atlantis_groups <- atlantis_groups %>% select(Code, Name, LongName)

race_species_all <- read.csv("../data/RACE_species_goa_Atlantis.csv", fileEncoding = "UTF-8-BOM")
race_species <- race_species_all[!is.na(race_species_all$Atlantis.group),] # drop NAs (egg cases, debris, etc.)
race_species <- race_species %>% select(Atlantis.group:Scientific.Name)

race_species <- race_species %>% left_join(atlantis_groups, by = c("Atlantis.group" = "Code"))
```

Read in AKFIN "Catch" data.
```{r}
catch <- read.csv("../data/race_catch_by_haul.csv", skip = 5)

# how many species do we have catch data for, from AKFIN?
length(levels(factor(catch$Species.Code))) # 1531

# and how many hauls?
length(levels(factor(catch$Haul.Join.ID))) # 12288
```

The Catch data does not include information about the effort. So let's use the "Haul Description" data set to obtain haul information. The column "Satisfactory Performance" will be used to subset these to the hauls that can be used.
```{r}
hauls_tmp <- read.csv("../data/Haul Descriptions.csv", fileEncoding = "UTF-8-BOM")
hauls <- hauls_tmp %>% 
  select(Haul.Join.ID, Distance.Fished..km., Net.Width..m., Satisfactory.Performance) %>%
  filter(Satisfactory.Performance == "Y") %>% 
  mutate(My.effort.km2 = Distance.Fished..km.* Net.Width..m. * 0.001) %>%
  select(Haul.Join.ID, My.effort.km2)
```

Now join the haul data to the catch data by Haul.Join.ID. Because now the haul data only contains the hauls we can use for effort/CPUE, base the join on that. Drop columns as appropriate. Then calculate new CPUE for weight and numbers from this. 
```{r}
catch_short <- catch %>% select(Year, Haul.Join.ID, Catch.Join.ID, Starting.Latitude..dd., Starting.Longitude..dd., Bottom.Depth, Species.Code, Weight..kg., Number.of.Fish)

catch_all <- hauls %>% left_join(catch_short) %>% mutate(cpue.kg.km2 = Weight..kg./My.effort.km2, cpue.ind.km2 = Number.of.Fish/My.effort.km2) %>%
  select(Year, Haul.Join.ID, Catch.Join.ID, Starting.Latitude..dd., Starting.Longitude..dd., Bottom.Depth, Species.Code, cpue.kg.km2, cpue.ind.km2)
```

However, some hauls in the haul data do not appear in the catch data. Discard those hauls that do not have information in the catch data. In addition, some records have no CPUE, biomass or numbers. Discard those too.
```{r}
catch_all <- catch_all %>% filter(!is.na(Starting.Latitude..dd.) & !is.na(cpue.kg.km2))
```

## Expand zeroes

Catch data only includes hauls with non-zero catch for each species. We need zero catches for sdmTMB.

Use only the catch data here, and see how it compares to the joined data frame - that should help spot issues. We can start from catch_all, which is a joined dataframe of catch and haul sets. 

The zero-inflated catch data is massive, so we need to query one species at a time. 
```{r}
# rename column names for compact code

catch_all <- catch_all %>% set_names(c("year","haul","catch","lat","lon","depth","species","biom","num"))

# add a column with Atlantis group code and name

key <- race_species %>% select(Species.Code, Atlantis.group, Name)
catch_all <- catch_all %>% left_join(key, by = c("species" = "Species.Code"))

catch_all <- catch_all %>% filter(!is.na(Name)) # discard those cases that do not map to Atlantis groups, for now

levels(factor(race_species$Name))
```

Have a look at the data.
```{r, fig.width = 12, fig.height = 16}
ggplot(data = catch_all, aes(x = year))+
  geom_bar(stat = "count")+
  theme_minimal()+
  facet_wrap(~Name, scales = "free_y", ncol = 5)+
  labs(title = "Hauls with non-zero catch")
```

And spatially.
```{r, fig.width = 10, fig.height=12}
catch_sf <- catch_all %>% st_as_sf(coords = c(x = "lon", y = "lat"), crs = "WGS84")
race_extent <- st_bbox(catch_sf) # set the extent of the race data

coast <- map("worldHires", regions = c("Canada", "USA"), plot = FALSE, fill = TRUE)
coast <- coast %>% st_as_sf()

group <- "Cod"

ggplot()+
  geom_sf(data = catch_sf[catch_sf$Name == group,], aes(color = log1p(biom)))+
  geom_sf(data = coast)+
  coord_sf(xlim = c(race_extent$xmin,race_extent$xmax),ylim = c(race_extent$ymin,race_extent$ymax))+
  scale_color_viridis()+
  theme_minimal()+
  facet_wrap(~year, ncol = 2)
```
See text to RACE.Rmd for thoughts about sample sizes of each group and what we should use (those refer to individuals, so for groups like corals and sponges there are fewer data points, but the main ideas hold).

Make a key that specify which groups we want to do this for. See RACE.Rmd for details. For now we are only doing: (1) invertebrate groups that we have no length-frequency for, eliminating the need for RACEBASE and K's code; and (2) aggregated vertebrate groups that we do not have from EFH, and for which the data is not too sparse (see RACE.Rmd).
```{r}
key1 <- levels(factor(race_species$Name))
do_sdm <- c(0,1,1,0,0,0,0,1,1,1,1,1,1,1,0,1,0,0,0,1,1,1,0,0,1,0,0,1,0,0,0,0,0,0,0,0,0,0,0,0,1,1,1,1,0,1,1,0,0,1,1,1,0)
key1 <- data.frame(key1, do_sdm)
```

Discard the species/groups that we will analyse elsewhere.
```{r}
catch_atlantis_tmp <- catch_all %>% filter(Name %in% key1[which(key1$do_sdm == 1),"key1"])
catch_atlantis <- catch_atlantis_tmp %>% group_by(year,haul,lat,lon,depth,Atlantis.group,Name) %>% summarise(biom = sum(biom), num = sum(num)) %>% ungroup()
```

Select which group we want to query.
```{r}
#group <- "Sablefish"
```

```{r}
# add up biomass and num cpue for Atlantis groups instead of species
#catch_atlantis <- catch_all %>% group_by(year,haul,lat,lon,depth,Atlantis.group,Name) %>% summarise(biom = sum(biom), num = sum(num)) %>% ungroup() # this step sums cpue of different species that belong to the same Atlantis group within the same haul - so the number of data points decreases for multi-species groups. Maps should stay the same though.

#length(levels(factor(catch_all$haul))) # check we have not lost hauls
#length(levels(factor(catch_atlantis$haul)))

# make a data frame with a unique row for each haul. This has to use all hauls

haul.ids <- catch_atlantis %>% select(year,haul,lat,lon,depth) %>% distinct() %>% arrange(haul)
group.ids <- catch_atlantis %>% select(Atlantis.group, Name) %>% distinct() %>% arrange(Atlantis.group) #%>% filter(Name == group)

long.hauls <- haul.ids[rep(seq_len(nrow(haul.ids)), each = nrow(group.ids)),]
long.groups <- group.ids[rep(seq_len(nrow(group.ids)), nrow(haul.ids)),]      

catch_zeroes <- data.frame(long.hauls, long.groups)
rm(long.hauls, long.groups)

test <- apply(catch_zeroes, 1, function(x) {ifelse(length(which(catch_atlantis$haul == x[2] & catch_atlantis$Atlantis.group == x[6]))>0,
                       catch_atlantis$biom[which(catch_atlantis$haul == x[2] & catch_atlantis$Atlantis.group == x[6])], 0)})

catch_zeroes <- catch_zeroes %>% 
  rowwise() %>% # this takes almost 10 mins
  mutate(biom = ifelse(length(which(catch_atlantis$haul == haul & catch_atlantis$Atlantis.group == Atlantis.group))>0,
                       catch_atlantis$biom[which(catch_atlantis$haul == haul & catch_atlantis$Atlantis.group == Atlantis.group)], 0),
         num = ifelse(length(which(catch_atlantis$haul == haul & catch_atlantis$Atlantis.group == Atlantis.group))>0,
                       catch_atlantis$num[which(catch_atlantis$haul == haul & catch_atlantis$Atlantis.group == Atlantis.group)], 0))
```

Some diagnostics follow. Do not run.
```{r}
# cpue <- read.csv("../data/race_cpue_by_haul.csv", skip = 5)
```

```{r}
# dog <- catch_zeroes %>% filter(Atlantis.group == "DOG")
# dog_cpue <- cpue %>% filter(Common.Name == "spiny dogfish")
# dog_cpue <- dog_cpue %>% select(Haul.Join.ID, Weight.CPUE..kg.km2., Number.CPUE..number.km2.)
# 
# dog <- dog %>% left_join(dog_cpue, by = c("haul" = "Haul.Join.ID")) 
# dog <- dog %>% mutate(biom_diff = biom - Weight.CPUE..kg.km2., num_diff = num - Number.CPUE..number.km2.)
# 
# par(mfrow = c(1,2))
# hist(dog$num_diff)
# hist(dog$biom_diff)
# 
# max(dog$num_diff, na.rm = TRUE)
# max(dog$biom_diff, na.rm = TRUE)
```
All good.

This data frame must have one row per tow. Let's plot it and compare with above. If I remove all zeroes, we should have left what we have above for single-species groups, but fewer data points for multi-species groups, since we have collapsed those inot one Atlantis group whereas in the Catch data there may have been multiple records for the same groups within a tow.
```{r, fig.width=12, fig.height=16}
# ggplot(data = catch_zeroes[catch_zeroes$biom>0,], aes(x = year))+
#   geom_bar(stat = "count")+
#   theme_minimal()+
#   labs(title = paste("Non-zero data points for", group, sep = " "))+
#   facet_wrap(~Name, scales = "free")
```

Save to RData, to open in sdmTMB. We probably forgot about some variables, but we will see.
```{r}
cpue_non_efh_groups <- catch_zeroes
save(cpue_non_efh_groups, file = "../output/cpue_non_efh_groups.Rdata")
```
